---
title: 'Visualizing International Space Station (ISS) Information via APIs and `leaflet`'
author: "Simon Marks"
date: "Updated: 4/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Open-Notify API

This was a lab completed for STAT 431 (Advanced Statistical Computing in R) in spring 2020. I updated my R markdown, as I was publicly publishing this small project on GitHub to showcase my familiarity with APIs via R and `leaflet`. I produce two interactive maps displaying information about ISS pass times (UTC) over the 51 capitals of the U.S states and territories:

1. Map shows includes the name of the state capital and the soonest predicted pass time as hover-over information for each marker. The click information includes the name of the state capital and the next three predicted pass times.

2. Map shows overlay lines connecting the U.S. state capitals in order of pass times (essentially east to west), as to show the expected pass order.

API used for this project: [link](http://open-notify.org/Open-Notify-API/ISS-Pass-Times/)

### Pass Times for U.S. State Captials

```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
library(lubridate)

# capitals data (lat , long)
capitals <- read_table2("https://people.sc.fsu.edu/~jburkardt/datasets/states/state_capitals_ll.txt", col_names = FALSE) %>% 
  rename(state = X1, latitude = X2, longitude = X3) %>% 
  # D.C and US are duplicate entries in the data set
  filter(!state %in% c("US"))

# data with names of capitals
capitals_names <-  read_delim("https://people.sc.fsu.edu/~jburkardt/datasets/states/state_capitals_name.txt", delim = "  ", col_names = FALSE) %>% 
  rename(state = X1, capital = X2) %>% 
  # D.C and US are duplicate entries in the data set
  filter(!state %in% c("US"))
```

```{r}
# Function to query API for next three pass times
Query <- function(state, latitude, longitude) {
  res <- GET("http://api.open-notify.org/iss-pass.json",
      query = list(lat = latitude, lon = longitude, n = 3))
  data_res <- fromJSON(rawToChar(res$content))
  pass_time_no <- seq(from = 1, to = nrow(data_res$response), by = 1)
  set <- cbind(pass_time_no, data_res$response) %>% 
    pivot_wider(names_from = pass_time_no, values_from = c(duration, risetime)) %>% 
    mutate(latitude = latitude) %>% 
    mutate(longitude = longitude) %>% 
    mutate(state = state)
  return(set)
}

# Returns a list of tibbles
PT_list <- capitals %>% 
  pmap(Query)

# Data frame for mapping, any missing obs due to data update will result in NA
PT_data <- bind_rows(PT_list) %>% 
  select(-contains("duration"))

# Times as POSIXct, join capital names
PT_data_dates <- PT_data %>% 
  mutate(risetime_1 = as_datetime(risetime_1)) %>% 
  mutate(risetime_2 = as_datetime(risetime_2)) %>% 
  mutate(risetime_3 = as_datetime(risetime_3)) %>% 
  inner_join(capitals_names, by = "state")
```

#### Data Preview
```{r}
head(PT_data, n = 10)
```

### Mapping the Data

```{r}
library(leaflet)

# custom icon
spaceicon <- makeIcon(
  iconUrl = "https://icons.iconarchive.com/icons/goodstuff-no-nonsense/free-space/256/astronaut-helmet-icon.png", iconWidth = 30, iconHeight = 30)

# Map with popup and labels containing pass time info
map1 <- leaflet(data = PT_data_dates) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -98.5795, lat = 39.8283, zoom = 3.2) %>% 
  setMaxBounds(lng1 = -157,
               lat1 = 14,
               lng2 = -58,
               lat2 = 64) %>% 
  addMarkers(lng = ~longitude, lat = ~latitude, icon = spaceicon,
             label = paste(PT_data_dates$capital, ":", PT_data_dates$risetime_1),
             popup = paste("Capital: ", PT_data_dates$capital, "<br>",
                           "Next 3 ISS Passes:", "<br>",
                           "1. ", PT_data_dates$risetime_1, "<br>", 
                           "2. ", PT_data_dates$risetime_2, "<br>", 
                           "3. ", PT_data_dates$risetime_3))
```

#### Pass Times (UTC) for the ISS: United States Capitals
```{r echo = FALSE}
map1
```

### Drawing the Route of the ISS

```{r}
# sort data for drawing ISS route, based on first pass time
PT_data_sort <- PT_data_dates %>% 
  arrange(risetime_1)

# Map from part 2, but with added polylines
map2 <- leaflet(data = PT_data_sort) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -98.5795, lat = 39.8283, zoom = 3.2) %>% 
  setMaxBounds(lng1 = -157,
               lat1 = 14,
               lng2 = -58,
               lat2 = 64) %>% 
  addPolylines(lng = ~longitude, lat = ~latitude, color = "blue") %>% 
  addMarkers(lng = ~longitude, lat = ~latitude, icon = spaceicon,
             label = paste(PT_data_sort$capital, ":", PT_data_sort$risetime_1),
             popup = paste("Capital: ", PT_data_sort$capital, "<br>",
                           "Next 3 ISS Passes:", "<br>",
                           "1. ", PT_data_sort$risetime_1, "<br>", 
                           "2. ", PT_data_sort$risetime_2, "<br>", 
                           "3. ", PT_data_sort$risetime_3))
```

#### Pass Times (UTC) for the ISS: United States Capitals
##### Route of 1st pass over each city is charted in blue 
```{r echo = FALSE}
map2
```